---
title: "PS270 Final Project"
author: "Harrison Crawford"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

I am interested in exploring data related to how differences in demographics correspond to mortality rates of various cancers. 

```{r}
library(tidyverse)
```

```{r}
finaldata <- read_csv("finaldata4.csv")
```
```{r}
print(finaldata)
```

```{r}
race_hospital_bar <- finaldata |>
  ggplot(aes(x = Race, y = `Distance to hospital, median (IQR), km`)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(race_hospital_bar)
```

The massive disparities here can partially be attributed to the fact that the vast majority of people sample were white; it may be more informative to compare percentages instead of absolute number. 

```{r}
finaldataperc <- read_csv("finaldataperc.csv")
```
It is notable that the percentages do not refer to the percentage of the race out of 100%; for example, the variable "Uninsured" does not contain 6% White, 9% Black, etc. Rather, it means that of the insurance sub-category (that includes the variables "Private insurance", "Medicaid or Medicare", and "Uninsured"), 6% of White people are uninsured, while 76% of White people have private insurance and 18% are on Medicaid or Medicare.

The sub-categories are:

- Deceased
- Cancer stage ("Stage I", "Stage II", "Stage III", "Stage IV", "NA (CNS)")
- Type of cancer ("Breast", "Cervical", "CNS", "Colon or rectal", "Lung", "Lymphoma", "Melanoma", "Ovarian", "Sarcoma", "Testicular")
- Gender ("Female", "Male")
- Distance to hospital, km (this is not a percentage)
- Income ("Higher income", "Lower income")
- Urban/rural ("Metropolitan", "Urban-rural")
- Education ("More education", "Less education")
- Insurance ("Private insurance", "Medicaid or Medicare", "Uninsured")
- Treatment received ("Surgery", "Chemotherapy", "Radiotherapy")


```{r}
finaldataperc
```


```{r}
race_gov_insure <- finaldataperc |>
  ggplot(aes(x = Race, y = `Medicaid or Medicare`)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(race_gov_insure)
```

## Milestone 4

```{r}
deceased_graph <- finaldataperc |>
  ggplot(aes(x = Race, y = Deceased)) + 
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(deceased_graph)
```

The graph above demonstrates the percentages of people included in the study within each race that died of cancer. Among those included in the study, 15% of white people died, 24% of Black people died, 20% of American Indians or Alaska Natives died, 16% of Asian people died, and 22% of Native Hawaiians or other Pacific Islanders died. Across all races, 16% of people who had cancer died from it. These numbers indicate that all races had a higher rate of death from cancer than the average of all people, with the exceptions of white people, who had a lower average, and Asian people, whose average equaled the total population. Since there is such a disparity between the death rates of white people and most other races, the data suggests that white people have a higher chance of surviving cancer across all cancer types.

The reason that the average is so low relative to the averages of all other races is because the subjects in this study were overwhelmingly white, bringing the average down considerably. It is the reason the data in the graph above looks at the percentages, rather than raw numbers, because the raw numbers would show that a larger number of white people died than any other race; this does not reflect the rate of death, it is reflective of how many white people were included in the study. 

Future data analysis can ascertain whether this trend remains consistent across each type of cancer. It is also worth analyzing whether other factors, such as rates of insurance coverage, education, and high income level follow the trends here: do white people have higher rates of being covered by private insurance? Getting higher education? Receiving more income? If there is a correlation, it is possible that those factors influence the rates at which people who have cancer die from it. 







