---
title: "PS270 Final Project"
author: "Harrison Crawford"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---


## Introduction

Cancer is an interesting and tragic condition due to its prevalence and mortality rate. It is also very diverse, with several different types of cancers and different rates of incidence among different populations, determined by environmental and genetic factors. It's this diversity in experience with cancer that naturally asks the question of how it is experienced across different groups of people. The research question of interest here is how the mortality rates of colon cancer differs across education, race, and income groups. As a hypothesis, it is expected that the mortality rate of cancer will be higher among those that are nonwhite and have lower incomes, with less education. This is because, of course, those with lower incomes likely have received less education and have less access to quality healthcare and therefore may not receive the treatment they need. There is also a correlation between race and income groups, with nonwhite people having a higher proportion of individuals living below the poverty line as compared to white people. In this analysis, we will analyze how the aforementioned three factors relate directly to mortality rates. We will also observe how factors such as race and income correlate with other factors that may more directly influence mortality rates, such as proportions of uninsured people and rates of late-stage diagnosis (Stage III or IV). This information is important because, upon discerning how mortality rates, and therefore healthcare disparities, are distributed across demographic groups, governments and organizations can more effectively create initiatives to correct these disparities by providing essential healthcare services to those who need it.   

The following data comes from the paper "Racial Disparities in Cancer Stage at Diagnosis and Survival for Adolescents and Young Adults" by Taparra, Kekumano, et al., published in the National Library of Medicine (https://pmc.ncbi.nlm.nih.gov/articles/PMC11365006/) on August 30, 2024. 

## Data

Loading the datasets and packages: 

```{r}
library(tidyverse)
finaldata <- read_csv("finaldata4.csv")
finaldataperc <- read_csv("finaldataperc.csv")
```

```{r}
knitr::kable(finaldata)
```


The study looks at five different races: 1) American Indian/Alaska Native, 2) Asian, 3) Black, 4) Native Hawaiian or Other Pacific Islander, and 5) White. There is also a sixth row to show the value of the variable across the entire population sample (Total). 

When looking at this study, there are two data sets to consider: the one labeled "finaldata" (finaldata4.csv), which contains the absolute numbers of peope who fit into each category in the study. There is also "finaldataperc" (finaldataperc.csv), which contains this same data in percentages. For the purposes of this analysis, finaldataperc will be the data set that is primarily used, for the following reason: 

```{r}
finaldata |>
  ggplot(aes(x = Race, y = `Medicaid or Medicare`)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The massive disparities here can partially be attributed to the fact that the vast majority of people sampled were white; therefore, it is informative to compare percentages instead of absolute number: 

```{r}
finaldataperc |>
  ggplot(aes(x = Race, y = `Medicaid or Medicare`)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


It is notable that the percentages do not refer to the percentage of the category, but percentage of the race; for example, the variable "Uninsured" does not contain 6% White, 9% Black, etc. Rather, it means that within the insurance sub-category (that includes the variables "Private insurance", "Medicaid or Medicare", and "Uninsured"), 76% of White people have private insurance, 18% are on Medicaid or Medicare, and 6% are uninsured.

The sub-categories are:

- Deceased
- Cancer stage ("Stage I", "Stage II", "Stage III", "Stage IV", "NA (CNS)")
- Type of cancer ("Breast", "Cervical", "CNS", "Colon or rectal", "Lung", "Lymphoma", "Melanoma", "Ovarian", "Sarcoma", "Testicular")
- Gender ("Female", "Male")
- Distance to hospital, km (this is not a percentage)
- Income ("Higher income", "Lower income")
- Urban/rural ("Metropolitan", "Urban-rural")
- Education ("More education", "Less education")
- Insurance ("Private insurance", "Medicaid or Medicare", "Uninsured")
- Treatment received ("Surgery", "Chemotherapy", "Radiotherapy")

For this analysis, we are only looking at the "Colon or rectal" cancer type. We are also not looking at the distance to hospital and urban/rural categories, simply because they are not of interest here.   



## Results

## Conclusion




# Project thoughts

I am interested in exploring data related to how differences in demographics correspond to mortality rates of various cancers. 


```{r}
print(finaldata)
```

```{r}
finaldataperc
```


```{r}
finaldataperc |>
  ggplot(aes(x = Race, y = `Medicaid or Medicare`)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Milestone 4

```{r}
deceased_graph <- finaldataperc |>
  ggplot(aes(x = Race, y = Deceased)) + 
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(deceased_graph)
```

The graph above demonstrates the percentages of people included in the study within each race that died of cancer. Among those included in the study, 15% of white people died, 24% of Black people died, 20% of American Indians or Alaska Natives died, 16% of Asian people died, and 22% of Native Hawaiians or other Pacific Islanders died. Across all races, 16% of people who had cancer died from it. These numbers indicate that all races had a higher rate of death from cancer than the average of all people, with the exceptions of white people, who had a lower average, and Asian people, whose average equaled the total population. Since there is such a disparity between the death rates of white people and most other races, the data suggests that white people have a higher chance of surviving cancer across all cancer types.

The reason that the average is so low relative to the averages of all other races is because the subjects in this study were overwhelmingly white, bringing the average down considerably. It is the reason the data in the graph above looks at the percentages, rather than raw numbers, because the raw numbers would show that a larger number of white people died than any other race; this does not reflect the rate of death, it is reflective of how many white people were included in the study. 

Future data analysis can ascertain whether this trend remains consistent across each type of cancer. It is also worth analyzing whether other factors, such as rates of insurance coverage, education, and high income level follow the trends here: do white people have higher rates of being covered by private insurance? Getting higher education? Receiving more income? If there is a correlation, it is possible that those factors influence the rates at which people who have cancer die from it. 







