---
title: "PS270 Final Project"
author: "Harrison Crawford"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

I am interested in exploring data related to how differences in demographics correspond to mortality rates of various cancers. 

```{r}
library(tidyverse)
```

```{r}
finaldata <- read_csv("finaldata4.csv")
```
```{r}
print(finaldata)
```

```{r}
race_hospital_bar <- finaldata |>
  ggplot(aes(x = Race, y = `Distance to hospital, median (IQR), km`)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(race_hospital_bar)
```

The massive disparities here can partially be attributed to the fact that the vast majority of people sample were white; it may be more informative to compare percentages instead of absolute number. 

```{r}
finaldataperc <- read_csv("finaldataperc.csv")
```
It is notable that the percentages do not refer to the percentage of the race out of 100%; for example, the variable "Uninsured" does not contain 6% White, 9% Black, etc. Rather, it means that of the insurance sub-category (that includes the variables "Private insurance", "Medicaid or Medicare", and "Uninsured"), 6% of White people are uninsured, while 76% of White people have private insurance and 18% are on Medicaid or Medicare.

The sub-categories are:

- Cancer stage ("Stage I", "Stage II", "Stage III", "Stage IV", "NA (CNS)")
- Type of cancer ("Breast", "Cervical", "CNS", "Colon or rectal", "Lung", "Lymphoma", "Melanoma", "Ovarian", "Sarcoma", "Testicular")
- Gender ("Female", "Male")
- Distance to hospital, km (this is not a percentage)
- Income ("Higher income", "Lower income")
- Urban/rural ("Metropolitan", "Urban-rural")
- Education ("More education", "Less education")
- Insurance ("Private insurance", "Medicaid or Medicare", "Uninsured")
- Treatment received ("Surgery", "Chemotherapy", "Radiotherapy")


```{r}
finaldataperc
```


```{r}
race_gov_insure <- finaldataperc |>
  ggplot(aes(x = Race, y = `Medicaid or Medicare`)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(race_gov_insure)
```


